---
title: Policy & Abortion EDA
author: Kayla Manning
output: 
  pdf_document:
    latex_engine: xelatex
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}

knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      echo = FALSE)

# loading libraries as needed

{
  # basic cleaning tasks
  
  library(tidyverse)
  library(readxl)
  library(janitor)
  
  # dealing with NA data
  
  library(naniar)
  
  # mixed effects modeling
  
  library(lme4)
  
  # for maps
  
  library(maps)
  library(gganimate)
  library(transformr)
}

# data <3

{
  # reading in gestational data that I scraped

  gestation <- read_csv('raw-data/scraped/gestational_combined.csv') %>% 
    select(-X1) %>% distinct()
  
  # looping through the different excel sheets to get the abortion count data
  
  {
    abortion_counts <- tibble()
    for (y in 2010:2019) {
      storage <- read_excel('raw-data/downloads/abortions_2010-2019.xls', 
                            skip = 1, sheet = as.character(y)) %>% 
        select(-1) %>% 
        clean_names() %>% 
        mutate(year = y) %>% 
        rename(state_location = state_area) %>% 
        select(state_location, year, everything())
      abortion_counts <- bind_rows(abortion_counts, storage)
    }
  }
  
  # reading in the birth & policy data
  
  births <- read.delim('raw-data/downloads/births_2007-2020.txt') %>% 
    clean_names() %>% 
    select(-c(notes, state_code, year_code))
  policies <- read_csv('raw-data/clean_policies.csv')
}

```

# Abortion data

First I will see which years the states did not fully report the data under the below criteria:

** Reporting area did not report, or only reported by occurrence; because abortion numbers by residence are available only from other reporting areas where residents obtained abortions, data by state/area of maternal residence should be interpreted with caution. 	

```{r didnt_report}

# I'm curious to see if all of the same states have "*" in every year

abortion_counts %>% 
  select(state_location, year) %>% 
  mutate(didnt_report = as.numeric(str_detect(state_location, fixed('**')))) %>% 
  group_by(state_location) %>% 
  summarise(nonreport_yrs = sum(didnt_report)) %>% 
  filter(nonreport_yrs != 0)

# want to get a list of years in which they didn't report... I should exclude
# these states from the analysis because they wouldn't have actually had 0
# abortions... it is an issue that CA didn't report though because they would
# likely have a high number since they're very democratic

nonreporting_states <- c('California', 'District of Columbia', 'Florida', 'Maryland', 'New Hampshire')
nonreport_years_df <- abortion_counts %>% 
  filter(str_detect(state_location, fixed('**'))) %>% 
  select(state_location, year) %>% 
  group_by(state_location) %>% 
  summarise(years = list(unique(year)))

nonreport_years_df %>% 
  filter(str_detect(state, nonreporting_states[1]))

```

Next, I check whether or not these asterisks correspond with states that have missing in-state data. That appears to be the case in most instances, except there are two years where DC did not report any in-state abortions and 10 years where Wyoming did not report any in-state abortions.

```{r clean_count}

# replacing missing data with NA values

abortion_counts <- abortion_counts %>% 
  mutate(across(everything(), function(x) na_if(x, '--')))

# converting df to long format

abortion_long <- abortion_counts %>% 
  pivot_longer(alabama:ncol(.), 
               names_to = 'state_residence', 
               values_to = 'count') %>% 
  mutate(state_residence = str_replace_all(state_residence, '_', ' ')) %>% 
  mutate(across(c(state_location, state_residence), str_to_title),
         count = as.integer(count)) %>% 
  mutate(across(c(state_location, state_residence), function(x) str_replace(x, ' Of ', ' of ')))

# checking to make sure missingness corresponds to '**' in the data

abortion_long %>% 
  mutate(state_location = str_remove_all(state_location, fixed('**'))) %>% 
  filter(state_location == state_residence,
         is.na(count)) %>% 
  group_by(state_location) %>% 
  count()

```

```{r out_of_state_counts}

# getting top out-of-state abortions, separated by year

abortion_long %>% 
  filter(state_location != state_residence,
         !str_detect(state_residence, 'Location'),
         !str_detect(state_residence, 'Residence'),
         !str_detect(state_location, 'Total')) %>% 
  filter(!(str_detect(state_location, 'New York') & 
             str_detect(state_residence, 'New York'))) %>% 
  arrange(desc(count))

# doing the same thing, but aggregated across all years in the data

abortion_long %>% 
  filter(state_location != state_residence,
         !str_detect(state_residence, 'Location'),
         !str_detect(state_residence, 'Residence'),
         !str_detect(state_location, 'Total')) %>% 
  filter(!(str_detect(state_location, 'New York') & 
             str_detect(state_residence, 'New York'))) %>% 
  group_by(state_location, state_residence) %>% 
  summarize(count = sum(count)) %>% 
  arrange(desc(count))

# also worth exploring top 10 by total by location of service

abortion_long %>% 
  filter(str_detect(state_residence, 'Total'),
         !str_detect(state_location, 'Total')) %>% 
  arrange(desc(count))

# aggregating top locations across all years

abortion_long %>% 
  filter(str_detect(state_residence, 'Total'),
         !str_detect(state_location, 'Total')) %>% 
  group_by(state_location) %>% 
  summarize(count = sum(count)) %>% 
  arrange(desc(count))

# I want states with highest abortion rates... dropping NAs in abortion counts
# for now

abortion_rates <- abortion_long %>% 
  inner_join(births, by = c('state_location' = 'state', 'year')) %>% 
  mutate(in_state = case_when(state_location == state_residence ~ 'in_state',
                              str_detect(state_residence, 'Total') ~ 'total',
                              TRUE ~ 'out_of_state')) %>% 
  group_by(state_location, year, in_state, births, total_population) %>% 
  summarise(abortions = sum(count, na.rm = TRUE)) %>% 
  summarize(abortion_per_1k_births = abortions / (births / 1000),
            abortion_per_100k_pop = abortions / (total_population / 100000),
            .groups = 'drop') %>% 
  select(-births)

# visualizing abortion rates per population & births... highly correlated. for
# substantive reasons (since birth counts include non-residents & my question
# has to do with abortions as a function of births rather than a function of
# residents) & to control for female population, I will do abortions per births

abortion_rates %>% 
  ggplot(aes(abortion_per_1k_births, abortion_per_100k_pop)) +
  geom_point() +
  geom_smooth(method = 'lm') +
  theme_minimal() +
  labs(title = 'Abortion rates as a function of population vs total births',
       subtitle = 
         paste('Correlation of', round(cor(abortion_rates$abortion_per_100k_pop,
                                           abortion_rates$abortion_per_1k_births), 3)))

```

# Policy data

```{r policy_counts}

# printing as-of dates for my policy years

policies %>% 
  select(year, as_of) %>% 
  distinct()

# getting counts of different policy types over the years

policies %>% 
  select(-c(state, as_of, accessed)) %>% 
  group_by(year, policy, level) %>% 
  summarize(n = n())

# doing the same but across all years... seeing most popular policies

policies %>% 
  select(-c(state, as_of, accessed)) %>% 
  group_by(policy, level) %>% 
  summarize(n = n(), .groups = 'drop') %>% 
  drop_na(level) %>% 
  arrange(desc(n))

# seeing how many enjoined policies there are... partial birth ban isn't really
# considered harsh since it's a very common ban on late term pregnancies, so I
# think it's okay to exclude enjoined observations... in addition, previous work
# by Blum et al. found that a parental notification statute passed with judicial
# bypass procedures had little impact on decision-making of minors obtaining
# abortions (i.e. I don't think having an enjoined policy, through a similar
# legal loophole, would be much different but I should consult with Henry or
# someone else about that once I get to writing)

policies %>% 
  select(-c(state, as_of, accessed)) %>% 
  filter(level == 'Enjoined') %>% 
  group_by(policy, level) %>% 
  summarize(n = n(), .groups = 'drop') %>% 
  arrange(desc(n))

```

# Abortions & births

```{r abortion_birth_graphs}

# checked and the only columns that didn't join were NYC (but NY state did) &
# 'Total by Residence'

abortion_births <- abortion_long %>% 
  mutate(state_location = str_remove_all(state_location, '[^A-Za-z0-9 ]'),
         state_location = str_replace_all(state_location, 'New York State', 'New York')) %>% 
  inner_join(select(births, state, year, births, total_population), 
             by = c('state_location' = 'state', 'year')) %>% 
  select(year, state_location, state_residence, count, births, total_population) %>% 
  rename(abortions = count)

# making some faceted plots

for (i in 1:5) {
  max_index <- i*10
  min_index <- max_index-9
  g <- abortion_births %>% 
    filter(state_residence == 'Total By Location of Service') %>% 
    drop_na(abortions) %>% 
    group_by(state_location, year) %>% 
    summarise(abortion_birth_ratio = sum(abortions) / sum(births),
              .groups = 'drop') %>% 
    filter(state_location %in% state.name[min_index:max_index]) %>% 
    ggplot(aes(year, abortion_birth_ratio)) +
    geom_point() +
    theme_classic() +
    labs(title = 'Abortion-to-birth ratio in the states over time',
         caption = 'Excluding NA observations',
         x = 'Year',
         y = 'Abortion rate') +
    scale_x_continuous(breaks = seq(2010, 2019, 2)) +
    geom_smooth() +
    facet_wrap(~state_location) +
    scale_y_continuous(limits = c(0, 0.4))
  print(g)
}

# want to see which state has had the greatest change in rate across the 9 year
# period

abortion_births %>% 
    filter(state_residence == 'Total By Location of Service') %>% 
    drop_na(abortions) %>% 
    group_by(state_location, year) %>% 
    summarise(abortion_birth_ratio = sum(abortions) / sum(births),
              .groups = 'drop') %>% 
  filter(year %in% c(2010, 2019)) %>% 
  pivot_wider(names_from = year, values_from = abortion_birth_ratio, names_prefix = 'x') %>% 
  mutate(change = x2019-x2010,
         state_location = fct_reorder(state_location, change)) %>% 
  drop_na(change) %>% 
  select(-c(x2010:x2019)) %>% 
  arrange(desc(change)) %>% 
  ggplot(aes(state_location, change)) +
  geom_col() +
  coord_flip() +
  labs(title = 'Change in abortion-to-birth ratio from 2010 to 2019',
       caption = 'Excludes states that did not report data for either of the two years',
       y = 'Change in abortion rate from 2010 to 2019',
       x = 'State of procedure')

```

# Gestational counts

```{r missing_gestat}

# there's a lot of missing data, so I should visualize missingness first

gestation
gg_miss_var(gestation)

# only 80% of the rows have matching total reported abortions and missing
# values... this indicates that there are many rows where there were negligible
# observations, so it is important to be mindful of this in any analysis

gestation  %>% 
  mutate(sum_all = select(., x0_13:x21) %>% rowSums(na.rm = TRUE),
         total_equals_sum = total_reported == sum_all * 1) %>% 
  pull(total_equals_sum) %>% 
  mean()

# checking proportion of observations in each category that are NA for each
# state (so not reported or a negligible amount were reported)

gestation %>% 
  pivot_longer(x0_13:x21, names_to = 'window') %>% 
  mutate(value = is.na(value)) %>% 
  group_by(state, window) %>% 
  summarize(prop_na = mean(value), .groups = 'drop') %>% 
  pivot_wider(names_from = window, values_from = prop_na)

# WY is excluded from every year except 2007 and 2019... what is the difference
# between NA and 0? it's because WY only reported for <=8 but had negligible
# counts for 9-13 and 14-15... for this reason I may just exclude WY from
# gestational calculations since the unknowns are split across pre- and post-13
# weeks

gestation %>% 
  filter(state == 'Wyoming')

```

```{r late_vs_early}

# now I want to get counts for pre13 and post13... since we have these missing
# counts, I will subtract total_report - x0_13 to get the post13 (which will
# hopefully pick up on the abortions that were excluded for having low counts in
# specific windows)

late_early_df <- gestation %>% 
  filter(state != 'Wyoming') %>% 
  mutate(post13 = total_reported - x0_13,
         prop_late = post13 / total_reported) %>% 
  select(state, prop_late, year) 

# across the board, there doesn't appear to be a difference, but what happens
# when we control for states

late_early_df %>% 
  ggplot(aes(as.factor(year), prop_late)) +
  geom_boxplot()

missing_gestation <- gestation %>% 
  pivot_longer(x0_13:total_reported, 
               names_to = 'category', 
               values_to = 'count') %>% 
  select(state, year, category, count) %>% 
  filter(is.na(count))

missing_gestation %>% 
  group_by(state, category) %>% 
  summarise(n = n()) %>% 
  arrange(category)

```

# Abortion & gestation combined

```{r joining_abortion_gestation}

# note that several observations are dropped if they don't report late-term abortions

abortion_births %>% 
  anti_join(late_early_df, by = c('year', 'state_location' = 'state'))

abortion_gestation_combined <- abortion_births %>% 
  inner_join(late_early_df, by = c('year', 'state_location' = 'state'))

```

# Gestation & policies combined

```{r policy_viz}

policy_scores <- read_csv('raw-data/policy_scores.csv')

policy_scores %>% 
  ggplot(aes(interstate_score, intrastate_score)) +
  geom_point() +
  labs(title = 'Intrastate score vs. interstate score (no clear relationship)') +
  geom_smooth()

policy_scores %>% 
  filter(state == 'Texas') %>% 
  pivot_longer(intrastate_score:interstate_score, names_to = 'type') %>% 
  ggplot(aes(year, value, color = type)) +
  geom_point() +
  labs(title = 'Intrastate / interstate scores vs. year in TX (no clear relationship)',
       subtitle = 'I really should not make this comparison because they are totally different scales') +
  geom_smooth()

# putting states in categories similar to Kaufman et al.

category_placements <- policy_scores %>% 
  pivot_longer(intrastate_score:interstate_score) %>% 
  group_by(name) %>% 
  mutate(first_cutoff = quantile(value, 1/3),
            second_cutoff = quantile(value, 2/3)) %>% 
  mutate(low = value <= first_cutoff,
         med = value > first_cutoff & value <= second_cutoff,
         high = value > second_cutoff) %>% 
  select(-c(first_cutoff, second_cutoff, value)) %>% 
  pivot_longer(low:high, names_to = 'category', values_to = 'placement') %>% 
  filter(placement) %>% 
  select(-placement) %>% 
  pivot_wider(names_from = name, values_from = category) %>% 
  mutate(within_between = paste(intrastate_score, interstate_score, sep = '-'),
         within_between = fct_relevel(within_between,
                                      levels = c('low-low', 'low-med', 'low-high',
                                                 'med-low', 'med-med', 'med-high',
                                                 'high-low', 'high-med', 'high-high'))) %>% 
  select(-c(intrastate_score, interstate_score))

# seeing overall breakdown between the categories

category_placements %>% 
  count(within_between)
category_placements %>% 
  count(within_between) %>% 
  ggplot(aes(within_between, n)) +
  geom_col() +
  labs(title = 'Substantial counts in each of the 9 policy categories',
       x = 'Intrastate-interstate score categories')

# making sure that states move around the 9 different categories

category_placements %>% 
  count(state, within_between) %>% 
  arrange(desc(n))

# checking out distribution by year

category_placements %>% 
  count(year, within_between) %>% 
  ggplot(aes(as_factor(year), n, fill = within_between)) +
  geom_col(position = 'dodge') +
  labs(title = 'Certain policy categories tend to dominate certain years')

category_placements %>% 
  count(year, within_between) %>% 
  ggplot(aes(within_between, n)) +
  coord_flip() +
  facet_wrap(~year) +
  geom_col(position = 'dodge') +
  labs(title = 'Overall, we go from a less strict landscape to more strict \nlandscape with time')

# want to visualize how it changes with time

category_placements %>% 
  count(year, within_between) %>% 
  arrange(desc(n))

category_placements %>% 
  count(year, within_between) %>% 
  ggplot(aes(within_between, n)) +
  geom_col() +
  transition_states(year) +
  labs(title = 'Policy category breakdown in {closest_state}')

```

```{r policy_maps}

# mapping the inter/intra scores on a map, animating by year
# loading US map data

states <- map_data('state')

map_policy_df <- policy_scores %>% 
  mutate(state = str_to_lower(state)) %>% 
  inner_join(states, by = c('state' = 'region'))

# making ggplot animations

map_policy_df %>%
  ggplot() +
  geom_polygon(aes(long, lat, group=group, fill = intrastate_score),
               color = 'black') +
  transition_states(year) +
  scale_fill_gradient2(low = 'blue',
                       mid = 'white',
                       high = 'red',
                       midpoint = 6,
                       labels = c(0, 3, 6, 9, 12)) +
  coord_fixed() +
  labs(title = "Within-state policy scores in {closest_state}")

map_policy_df %>%
  ggplot() +
  geom_polygon(aes(long, lat, group=group, fill = interstate_score),
               color = 'black') +
  transition_states(year) +
  scale_fill_gradient2(low = 'blue',
                       mid = 'white',
                       high = 'red') +
  coord_fixed() +
  labs(title = "Interstate policy scores in {closest_state}")

# now just doing intrastate for 2010 and 2019 side-by-side

map_policy_df %>% 
  filter(year %in% c(2010, 2019)) %>% 
  ggplot() +
  geom_polygon(aes(long, lat, group=group, fill = intrastate_score),
               color = 'black') +
  scale_fill_gradient2(high = 'red') +
  coord_fixed() +
  facet_wrap(~year, nrow = 2)+
  labs(title = 'Intrastate scores for 2010 and 2019')

# seeing the change in hostility

map_policy_df %>% 
  filter(year %in% c(2010, 2019)) %>% 
  select(-interstate_score) %>% 
  pivot_wider(names_from = year, values_from = intrastate_score, names_prefix = 'x') %>% 
  mutate(intra_diff = x2019-x2010) %>% 
  ggplot() +
  geom_polygon(aes(long, lat, group=group, fill = intra_diff),
               color = 'black') +
  scale_fill_gradient2(high = 'red3') +
  coord_fixed() +
  labs(title = 'Change in intrastate score from 2010 and 2019') +
  theme(panel.background = element_blank(),
        axis.text = element_blank(),
        axis.line = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank())

map_policy_df %>% 
  filter(year %in% c(2010, 2019)) %>% 
  select(-intrastate_score) %>% 
  pivot_wider(names_from = year, values_from = interstate_score, names_prefix = 'x') %>% 
  mutate(inter_diff = x2019-x2010) %>% 
  ggplot() +
  geom_polygon(aes(long, lat, group=group, fill = inter_diff),
               color = 'black') +
  scale_fill_gradient2(high = 'red3') +
  coord_fixed() +
  labs(title = 'Change in interstate score from 2010 and 2019') +
  theme(panel.background = element_blank(),
        axis.text = element_blank(),
        axis.line = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank())

```

```{r play_mods}

combine_all <- policy_scores %>% 
  mutate(state = ifelse(state == 'DC',
                        'District of Columbia',
                        state)) %>% 
  inner_join(abortion_gestation_combined, by = c('year', 'state' = 'state_location')) %>% 
  mutate(across(where(is.character), addNA))

library(lme4)
library(lmerTest)

combine_all_late <- combine_all %>% 
  select(-c(state_residence, total_population)) %>% 
  distinct() %>% 
  mutate(abortions_per_thousand_births = abortions / (births * 1000))

# fitting a full linear model, even though state and intrastate score are highly correlated

summary(full_lm <- glm(prop_late ~ interstate_score + intrastate_score + 
                        abortions_per_thousand_births,
                       family = binomial,
                       data = combine_all_late))
car::vif(full_lm)

full_glmer <- glmer(prop_late ~ interstate_score + intrastate_score + 
                            (1|year) + (1|state),
                          data = combine_all_late)
fixef(full_glmer)
summary(full_glmer)

```
