---
title: Policy & Abortion EDA
author: Kayla Manning
output: 
  pdf_document:
    latex_engine: xelatex
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}

knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      echo = FALSE)

# loading libraries as needed

{
  # basic cleaning tasks
  
  library(tidyverse)
  library(readxl)
  library(janitor)
  library(zoo)
  
  # dealing with NA data
  
  library(naniar)
  
  # mixed effects modeling
  
  library(lme4)
    
  # helping format tables
  
  library(reshape2)
  library(kableExtra)
  
  # for maps
  
  library(maps)
  library(gganimate)
  library(transformr)
}

# data <3

{
  # reading in gestational data that I scraped

  gestation <- read_csv('raw-data/scraped/gestational_combined.csv') %>% 
    select(-X1) %>% distinct()
  
  # looping through the different excel sheets to get the abortion count data
  
  {
    abortion_counts <- tibble()
    for (y in 2010:2019) {
      storage <- read_excel('raw-data/downloads/abortions_2010-2019.xls', 
                            skip = 1, sheet = as.character(y)) %>% 
        select(-1) %>% 
        clean_names() %>% 
        mutate(year = y) %>% 
        rename(state_location = state_area) %>% 
        select(state_location, year, everything())
      abortion_counts <- bind_rows(abortion_counts, storage)
    }
  }
  
  # reading in the birth & policy data
  
  births <- read.delim('raw-data/downloads/births_2007-2020.txt') %>% 
    clean_names() %>% 
    select(-c(notes, state_code, year_code))
  policies <- read_csv('raw-data/clean_policies.csv')
  policy_scores <- read_csv('raw-data/policy_scores.csv')
}

```

```{r abortion_long}

# replacing missing data with NA values

abortion_counts <- abortion_counts %>% 
  mutate(across(everything(), function(x) na_if(x, '--')))

# converting df to long format

abortion_long <- abortion_counts %>% 
  pivot_longer(alabama:ncol(.), 
               names_to = 'state_residence', 
               values_to = 'count') %>% 
  mutate(state_residence = str_replace_all(state_residence, '_', ' ')) %>% 
  mutate(across(c(state_location, state_residence), str_to_title),
         count = as.integer(count)) %>% 
  mutate(across(c(state_location, state_residence), 
                function(x) str_replace(x, ' Of ', ' of ')))

```

# Data

## Limitations

### Missing abortion data

#### Missing abortion counts

```{r missing_state_yrs_abortions}

# want to get a list of years in which they didn't report... I should exclude
# these states from the analysis because they wouldn't have actually had 0
# abortions... it is an issue that CA didn't report though because they would
# likely have a high number since they're very democratic

nonreporting_states <- c('California', 'District of Columbia', 
                         'Florida', 'Maryland', 'New Hampshire')
nonreport_years_df <- abortion_counts %>% 
  filter(str_detect(state_location, fixed('**'))) %>% 
  select(state_location, year) %>% 
  group_by(state_location) %>% 
  summarise(years = list(unique(year)))

missing_states <- sapply(nonreport_years_df$state_location, 
       function (x) str_replace_all(str_to_lower(str_remove_all(x, 
                                                                fixed('*'))), 
                                    ' ', '_'))

# getting a vector of each year that the data is missing for the respective state

{
  i <- 1
  for (state in missing_states) {
    vector_name <- paste0('missingyrs_', state)
    assign(vector_name, nonreport_years_df$years[i][[1]])
    i <- i+1
  }
}

# I will want to put this info in a table

missing_state_yr_table <- tibble()
for (i in 1:length(missing_states)) {
  this_state <- cbind(rep(missing_states[i], length(nonreport_years_df$years[[i]])),
      nonreport_years_df$years[[i]]) %>% 
  as.data.frame()
  missing_state_yr_table <- rbind(missing_state_yr_table, this_state)
}

# formatting the table

missing_state_yr_table %>% 
  rename(state = V1, year = V2) %>% 
  dcast(state ~ year) %>% 
  mutate(across(2:ncol(.), function(x) ifelse(is.na(x), '', 'X'))) %>% 
  mutate(state = str_replace(str_to_title(str_replace_all(state, '_', ' ')),
                             ' Of ', ' of ')) %>% 
  rename(State = state) %>% 
  kable(caption = 'States that do not report their abortion counts to CDC',
        booktabs = TRUE)


```

#### Missing gestational data

```{r missing_gestation_breakdown}

# checking proportion of observations in each category that are NA for each
# state (so not reported or a negligible amount were reported)

gestation %>% 
  pivot_longer(x0_13:x21, names_to = 'window') %>% 
  mutate(value = is.na(value)) %>% 
  group_by(state, window) %>% 
  summarize(prop_na = mean(value), .groups = 'drop') %>% 
  pivot_wider(names_from = window, values_from = prop_na)

# WY is excluded from every year except 2007 and 2019... what is the difference
# between NA and 0? it's because WY only reported for <=8 but had negligible
# counts for 9-13 and 14-15... for this reason I may just exclude WY from
# gestational calculations since the unknowns are split across pre- and post-13
# weeks

gestation %>% 
  filter(state == 'Wyoming')

# creating a table similar to the one I created for gestational data

# getting states that are only missing certain years

missing_some_gestation <- gestation %>% 
  select(state, year) %>% 
  group_by(state) %>% 
  count() %>% 
  filter(n != 13) %>% 
  pull(state)

# creating table for all missing

missing_all_df <- tibble(state = rep(missing_all_gestation, each = length(2007:2019)),
       year = rep(2007:2019, times = length(missing_all_gestation))) %>% 
  dcast(state ~ year) %>% 
  mutate(across(`2007`:`2019`, function(x) ifelse(is.na(x), '', 'X')))

# creating table for some missing & then joining

gestation %>% 
  filter(state %in% missing_some_gestation) %>% 
  dcast(state ~ year) %>% 
  mutate(across(`2007`:`2019`, function(x) ifelse(is.na(x), 'X', ''))) %>% 
  bind_rows(missing_all_df) %>% 
  arrange(state) %>% 
  rename(State = state) %>% 
  kable(caption = 'States that do not report gestational data to CDC',
        booktabs = TRUE)

```

### Issues with policy data

#### Policy timelines

```{r policy_dates}

# printing as-of dates for my policy years

policies %>% 
  select(year, as_of) %>% 
  distinct() %>% 
  rename(Year = year, `Earliest available policies` = as_of) %>% 
  kable(booktabs = TRUE,
        align = 'c',
        caption = 'Earliest available policy data on the Internet Archive') %>% 
  kable_styling(full_width = TRUE)

```

#### Enjoined policies

```{r policy_counts}

# seeing how many enjoined policies there are... partial birth ban isn't really
# considered harsh since it's a very common ban on late term pregnancies, so I
# think it's okay to exclude enjoined observations... in addition, previous work
# by Blum et al. found that a parental notification statute passed with judicial
# bypass procedures had little impact on decision-making of minors obtaining
# abortions (i.e. I don't think having an enjoined policy, through a similar
# legal loophole, would be much different but I should consult with Henry or
# someone else about that once I get to writing)

policies %>% 
  select(-c(state, as_of, accessed)) %>% 
  filter(level == 'Enjoined') %>% 
  group_by(policy, level) %>% 
  summarize(n = n(), .groups = 'drop') %>% 
  arrange(desc(n))

```

# Defining variables

## Policy scores

```{r score_visualizations}

policy_scores

```

```{r category_df}

# putting states in categories similar to Kaufman et al.

category_placements <- policy_scores %>% 
  pivot_longer(intrastate_score:interstate_score) %>% 
  group_by(name) %>% 
  mutate(first_cutoff = quantile(value, 1/3),
         second_cutoff = quantile(value, 2/3)) %>% 
  mutate(low = value <= first_cutoff,
         med = value > first_cutoff & value <= second_cutoff,
         high = value > second_cutoff) %>% 
  select(-c(first_cutoff, second_cutoff, value)) %>% 
  pivot_longer(low:high, names_to = 'category', values_to = 'placement') %>% 
  filter(placement) %>% 
  select(-placement) %>% 
  pivot_wider(names_from = name, values_from = category) %>% 
  mutate(within_between = paste(intrastate_score, interstate_score, sep = '-'),
         within_between = fct_relevel(within_between,
                                      levels = c('low-low', 'low-med', 'low-high',
                                                 'med-low', 'med-med', 'med-high',
                                                 'high-low', 'high-med', 'high-high'))) %>% 
  select(-c(intrastate_score, interstate_score))

```

```{r category_viz}

# seeing overall breakdown between the categories

category_placements %>% 
  count(within_between)
category_placements %>% 
  count(within_between) %>% 
  ggplot(aes(within_between, n)) +
  geom_col() +
  labs(title = 'Substantial counts in each of the 9 policy categories',
       x = 'Intrastate-interstate score categories') +
  theme_minimal()


```


## Dependent variables

### Import-export ratio

```{r ie_ratio}

# getting counts of non-resident abortions obtained within X state (i.e.
# abortions imported from other states)

imports <- abortion_long %>% 
  filter(state_location != state_residence) %>% 
  group_by(state_location, year) %>% 
  summarize(imports = sum(count, na.rm = TRUE),
            .groups = 'drop')

# getting counts of abortions obtained by residents of X state in OTHER states
# (i.e. abortions exported to other states)

exports <- abortion_long %>% 
  filter(state_location != state_residence) %>% 
  group_by(state_residence, year) %>% 
  summarize(exports = sum(count, na.rm = TRUE),
            .groups = 'drop')

# computing import/export ratio for each state & year, with import and export
# defined above

ie_df <- inner_join(imports, exports, 
           by = c('state_location' = 'state_residence', 
                  'year')) %>% 
  mutate(ie_ratio = imports / exports)

# plotting imports vs exports

ie_df %>% 
  ggplot(aes(exports, imports)) +
  geom_point() +
  labs(title = 'States that import more abortions also tend to export more abortions',
       subtitle = 
         paste('Correlation of', round(cor(ie_df$imports, ie_df$exports), 3))) +
  theme_minimal() +
  geom_smooth(method = 'lm')

# taking a look at the histogram... incredibly skewed

ggplot(ie_df, aes(ie_ratio)) +
  geom_histogram() +
  labs(title = 'Import / export ratio',
       subtitle = 'Taking the ratio standardizes the measure across states with different \npopulations and birthrates') +
  theme_minimal()

ggplot(ie_df, aes(log(ie_ratio))) +
  geom_histogram() +
  labs(title = 'log(Import / export ratio)',
       subtitle = 'Taking the ratio standardizes the measure across states with different \npopulations and birthrates') +
  theme_minimal()

```

### Abortion rates

Do I want all abortions obtained within the state I'm assuming?

```{r rates_metrics}

# I want states with highest abortion rates... dropping NAs in abortion counts
# for now

abortion_rates <- abortion_long %>% 
  inner_join(births, by = c('state_location' = 'state', 'year')) %>% 
  # mutate(in_state = case_when(state_location == state_residence ~ 'in_state',
  #                             str_detect(state_residence, 'Total') ~ 'total',
  #                             TRUE ~ 'out_of_state')) %>% 
  group_by(state_location, year, births, total_population) %>% 
  summarise(abortions = sum(count, na.rm = TRUE)) %>% 
  summarize(abortion_per_1k_births = abortions / (births / 1000),
            abortion_per_100k_pop = abortions / (total_population / 100000),
            .groups = 'drop') %>% 
  select(-births)

# visualizing abortion rates per population & births... highly correlated. for
# substantive reasons (since birth counts include non-residents & my question
# has to do with abortions as a function of births rather than a function of
# residents) & to control for female population, I will do abortions per births

abortion_rates %>% 
  ggplot(aes(abortion_per_1k_births, abortion_per_100k_pop)) +
  geom_point() +
  geom_smooth(method = 'lm') +
  theme_minimal() +
  labs(title = 'Abortion rates as a function of population vs total births',
       subtitle = 
         paste('Correlation of', round(cor(abortion_rates$abortion_per_100k_pop,
                                           abortion_rates$abortion_per_1k_births), 3)))

```

### Later-term abortions

```{r late_vs_early}

# now I want to get counts for pre13 and post13... since we have these missing
# counts, I will subtract total_report - x0_13 to get the post13 (which will
# hopefully pick up on the abortions that were excluded for having low counts in
# specific windows)

late_early_df <- gestation %>% 
  filter(state != 'Wyoming') %>% 
  mutate(post13 = total_reported - x0_13,
         prop_late = post13 / total_reported) %>% 
  rename(pre13 = x0_13) %>% 
  select(state, year, pre13, post13) 

```

```{r combine_dep_vars}

inner_join(ie_df, abortion_rates, by = c('state_location', 'year')) %>% 
  inner_join(late_early_df, by = c('state_location' = 'state', 'year'))

```


# Exploring the data

## Policies

```{r explore_policy_levels}

# getting counts of different policy types over the years... showing 10 that had
# the largest magnitudes of change over the period from 2010-2019

policies %>% 
  select(-c(state, as_of, accessed)) %>% 
  group_by(year, policy, level) %>% 
  summarize(n = n(), .groups = 'drop') %>% 
  filter(year %in% 2010:2019) %>% 
  pivot_wider(names_from = year, values_from = n) %>% 
  mutate(across(3:ncol(.), function(x) replace_na(x, 0))) %>% 
  arrange(desc(abs(`2019`-`2010`)))

# getting a count of most popular policy levels when aggregated across all
# states and years

policies %>% 
  select(-c(state, as_of, accessed)) %>% 
  group_by(policy, level) %>% 
  summarize(n = n(), .groups = 'drop') %>% 
  drop_na(level) %>% 
  arrange(desc(n)) %>% head(10)

```

## Out-of-state abortions

```{r out_of_state_counts}

# getting top out-of-state abortions, separated by year

abortion_long %>% 
  filter(state_location != state_residence,
         !str_detect(state_residence, 'Location'),
         !str_detect(state_residence, 'Residence'),
         !str_detect(state_location, 'Total')) %>% 
  filter(!(str_detect(state_location, 'New York') & 
             str_detect(state_residence, 'New York'))) %>% 
  arrange(desc(count))

# doing the same thing, but aggregated across all years in the data

summarize <- dplyr::summarize
abortion_long %>% 
  filter(state_location != state_residence,
         !str_detect(state_residence, 'Location'),
         !str_detect(state_residence, 'Residence'),
         !str_detect(state_location, 'Total')) %>% 
  filter(!(str_detect(state_location, 'New York') & 
             str_detect(state_residence, 'New York'))) %>% 
  group_by(state_location, state_residence) %>% 
  summarize(count = sum(count)) %>% 
  arrange(desc(count))

# also worth exploring top 10 by total by location of service

abortion_long %>% 
  filter(str_detect(state_residence, 'Total'),
         !str_detect(state_location, 'Total')) %>% 
  arrange(desc(count))

# aggregating top locations across all years

abortion_long %>% 
  filter(str_detect(state_residence, 'Total'),
         !str_detect(state_location, 'Total')) %>% 
  group_by(state_location) %>% 
  summarize(count = sum(count)) %>% 
  arrange(desc(count))

# out-of-state to in-state ratios

abortion_long %>% 
  mutate(across(c(state_residence, state_location), 
                function (x) str_remove_all(x, fixed('*')))) %>% 
  mutate(same_state = state_location == state_residence) %>% 
  group_by(state_location, same_state) %>% 
  summarise(aggregated = sum(count, na.rm = TRUE), .groups = 'drop') %>% 
  filter(!is.na(same_state)) %>% 
  pivot_wider(names_from = same_state, values_from = aggregated,
              names_prefix = 'instate') %>% 
  mutate(out_in_ratio = instateFALSE / instateTRUE) %>% 
  arrange(desc(out_in_ratio))

```

## Abortion rates

```{r abortion_birth_graphs}

# checked and the only columns that didn't join were NYC (but NY state did) &
# 'Total by Residence'

abortion_births <- abortion_long %>% 
  mutate(state_location = str_remove_all(state_location, '[^A-Za-z0-9 ]'),
         state_location = str_replace_all(state_location, 'New York State', 'New York')) %>% 
  inner_join(select(births, state, year, births, total_population), 
             by = c('state_location' = 'state', 'year')) %>% 
  select(year, state_location, state_residence, count, births, total_population) %>% 
  rename(abortions = count)

# making some faceted plots

for (i in 1:5) {
  max_index <- i*10
  min_index <- max_index-9
  g <- abortion_births %>% 
    filter(state_residence == 'Total By Location of Service') %>% 
    drop_na(abortions) %>% 
    group_by(state_location, year) %>% 
    summarise(abortion_birth_ratio = sum(abortions) / sum(births),
              .groups = 'drop') %>% 
    filter(state_location %in% state.name[min_index:max_index]) %>% 
    ggplot(aes(year, abortion_birth_ratio)) +
    geom_point() +
    theme_classic() +
    labs(title = 'Abortion-to-birth ratio in the states over time',
         caption = 'Excluding NA observations',
         x = 'Year',
         y = 'Abortion rate') +
    scale_x_continuous(breaks = seq(2010, 2019, 2)) +
    geom_smooth() +
    facet_wrap(~state_location) +
    scale_y_continuous(limits = c(0, 0.4))
  print(g)
}

# want to see which state has had the greatest change in rate across the 9 year
# period

abortion_births %>% 
    filter(state_residence == 'Total By Location of Service') %>% 
    drop_na(abortions) %>% 
    group_by(state_location, year) %>% 
    summarise(abortion_birth_ratio = sum(abortions) / sum(births),
              .groups = 'drop') %>% 
  filter(year %in% c(2010, 2019)) %>% 
  pivot_wider(names_from = year, values_from = abortion_birth_ratio, names_prefix = 'x') %>% 
  mutate(change = x2019-x2010,
         state_location = fct_reorder(state_location, change)) %>% 
  drop_na(change) %>% 
  select(-c(x2010:x2019)) %>% 
  arrange(desc(change)) %>% 
  ggplot(aes(state_location, change)) +
  geom_col() +
  coord_flip() +
  labs(title = 'Change in abortion-to-birth ratio from 2010 to 2019',
       caption = 'Excludes states that did not report data for either of the two years',
       y = 'Change in abortion rate from 2010 to 2019',
       x = 'State of procedure')

```
